---
title: "Summary of Storm Data from 1950 to 2011 and its Costs"
author: "Jeffrey Norton, Ph.D."
date: "March 10, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Original Data

The original data is [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and the National Weather Service Storm Data Documentation is [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).
The events in the database span from 1950 and end in November 2011. The data from earlier years is sparse with fewer
recorded events.  The data from more recent years is considered more complete.

# Data Processing

[Download](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and unzip the CSV file.  Read the table into a data table representing the raw storm data.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
StormData <- read.csv(file="repdata%2Fdata%2FStormData.csv", header=TRUE)
```

## Data Analysis

The steps taken in the data analysis are the following:

1. Determine and summarize the worst storm type for fatalities and injuries for all the United States from 1950 to 2011.

2. Determine and summarize the worst storm type for property and crop damage in dollars for all the Unites States from 1950 to 2011.

3. For each state (which includes US States, marine regions, etc.), determine a maximum of the three most costly storms with respect to fatalities, injuries, property, and crop damage.

# Results

## Fatalities and Injuries

Determine the worst Storm Type for fatalities and injuries across the United States.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
# For ease, create named vectors of the fatalities/injuries.
evtype <- unique(StormData$EVTYPE)
event.fatalities <- integer(length=length(evtype))
event.injuries   <- integer(length=length(evtype))
i <- 1
for( event in evtype ) {
  s <- sum(StormData[StormData$EVTYPE==event,]$FATALITIES)
  event.fatalities[i] <- s
  s <- sum(StormData[StormData$EVTYPE==event,]$INJURIES)
  event.injuries[i] <- s
  i <- i+1
}
names(event.fatalities) <- unique(StormData$EVTYPE)
names(event.injuries) <- unique(StormData$EVTYPE)

# Get max fatalities and injuries...
worst.f <- which(event.fatalities==max(event.fatalities))
worst.i <- which(event.injuries==max(event.injuries))
worst.fatal.name <- names(event.fatalities[worst.f])
worst.injury.name <- names(event.injuries[worst.i])
print(sprintf("Event and Count: Worst Fatalities: %s=%d.  Worst Injuries: %s=%d",
      worst.fatal.name, as.integer(event.fatalities[worst.f]),
      worst.injury.name, as.integer(event.injuries[worst.i])))

```

The worst storm type is `r worst.fatal.name` with `r prettyNum(as.integer(event.fatalities[worst.f]),scientific=FALSE,big.mark=",")` fatalities
and `r worst.injury.name` with `r prettyNum(as.integer(event.injuries[worst.i]),scientific=FALSE,big.mark=",")` injuries (total from 1950 to 2011).

The following plot shows injuries and fatalities across the United States for `r worst.fatal.name` events.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
# Add a year column to StormData (since we will use it three times...)
StormData[["YEARY"]] <-
  as.numeric(format(as.Date(vapply(strsplit(as.character(StormData$BGN_DATE), split=" "),
                                   '[', 1, FUN.VALUE=character(1)), format="%m/%d/%Y"), "%Y"))
# Create a data.frame with the year and the sum of each years fatalities...
unique.years <- unique(StormData[["YEARY"]])
fatality.sum <- injury.sum <- numeric(length(unique.years))
i <- 1
for( y in unique.years ) {
  fatality.sum[i] = sum(StormData[StormData$YEARY==y,]$FATALITIES)
  injury.sum[i] = sum(StormData[StormData$YEARY==y,]$INJURIES)
  i <- i+1
}
plot.df <- data.frame(unique.years, fatality.sum, injury.sum)
colnames(plot.df) <- c("year", "fatalities", "injuries")
# Now create a ggplot using the data frame
library(ggplot2)
library(tidyr)
print(
  plot.df %>%
    gather(Legend,Count, fatalities, injuries) %>%
    ggplot(aes(x=year,y=Count,color=Legend)) + geom_point() +
    labs(x="Year") + ggtitle("Fatalities and Injuries due to Storm Events in the United States"))
```

## Storm damage (property and crop)

Determine the worst Storm Type for property and crop damage in dollars across the United States.

### Interpreting Property and Crop Damage Amounts

```{r echo=TRUE, eval=TRUE, cache=TRUE}
# Calculate property and crop damage.  We need to set up an exponent table
# because the exponents are crazy for these entries:
# K M   B m + 0 5 6 ? 4 2 3 h 7 H - 1 8 ? 0 2 B k K m M
# Some make absolutely no sense, so we have to guess them.
# The documentation is of no use for the strange entries.
available.exponents <- unique(c(as.character(unique(StormData$CROPDMGEXP)),
                                as.character(unique(StormData$PROPDMGEXP))))
writeLines(sprintf("All used exponents for cost values:\n%s",
                   paste(available.exponents, collapse=", ")))
# The following list includes all the above plus an extra small "b" for billions.
exponent.lookup <- list(K=3, M=6, "1"=1, "2"=2, "3"=3, "4"=4, "5"=5,
                        "6"=6, "7"=7, "8"=8, "9"=9, B=9, m=6,
                        "0"=1, "+"=1, "-"=1, k=3, b=9, "?"=1, h=3, H=3)
# Get something printable with kable
exponent.lookup.df <- data.frame(unlist(exponent.lookup))
colnames(exponent.lookup.df) <- c("Exponent Value")
```

Amounts are given as a number and an exponent.  Unfortunately, exponents are not given in a clear and consistent
manner in many cases.  We have defined a lookup table for exponents of property and crop damage.  While many of the
values are defined in the [documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) or are obvious, many are not.  For example, what does "H" or "-" represent as exponents.  In cases where a value cannot be determined, exponent values
are set to "1".  All values for exponents as used in further calculations
and given here:
`r knitr::kable(exponent.lookup.df)`

Given these assumptions, calculate property and cost damage.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
l <- as.character(StormData$PROPDMGEXP)
l[which(l=="")] <- "1"
for( e in unique(l) ) {
  l[which(l==e)] <- exponent.lookup[[e]]
}
prop.damage <- StormData$PROPDMG * 10^(as.integer(l))
l <- as.character(StormData$CROPDMGEXP)
l[which(l=="")] <- "1"
for( e in unique(l) ) {
  l[which(l==e)] <- exponent.lookup[[e]]
}
crop.damage <- StormData$CROPDMG * 10^(as.integer(l))

# Store the property and crop damage values into the StormData data.frame as
# we will be using it from here on out.
StormData[["PROPDMGC"]] <- prop.damage
StormData[["CROPDMGC"]] <- crop.damage

event.prop.damage <- integer(length=length(evtype))
event.crop.damage <- integer(length=length(evtype))
i <- 1
for( event in evtype ) {
  s <- sum(StormData[StormData$EVTYPE==event,]$PROPDMGC)
  event.prop.damage[i] <- s
  s <- sum(StormData[StormData$EVTYPE==event,]$CROPDMGC)
  event.crop.damage[i] <- s
  i <- i+1
}
names(event.prop.damage) <- unique(StormData$EVTYPE)
names(event.crop.damage) <- unique(StormData$EVTYPE)

# Get max property damage...
worst.p <- which(event.prop.damage==max(event.prop.damage))
worst.c <- which(event.crop.damage==max(event.crop.damage))
worst.p.name <- names(event.prop.damage[worst.p])
worst.c.name <- names(event.crop.damage[worst.c])
writeLines(sprintf(
  "Event and Count:\nWorst Property Damage: %s=$%15.2f.\nWorst Crop Damage: %s=$%15.2f",
      worst.p.name, as.numeric(event.prop.damage[worst.p]),
      worst.c.name, as.numeric(event.crop.damage[worst.c])))

```
For property damage, the worst storm type is `r worst.p.name` with
`r prettyNum(as.numeric(event.prop.damage[worst.p]),scientific=FALSE,big.mark=",")` dollars damage.
For crop damage, the worst storm type is `r worst.c.name` with
`r prettyNum(as.numeric(event.crop.damage[worst.c]),scientific=FALSE,big.mark=",")` dollars damage.
Both values are the sum of all events from 1950 to 2011.

The following plot shows property damage across the United States for `r worst.p.name` events.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
Damage <- numeric(length(unique.years))
i <- 1

for( y in unique.years ) {
  Damage[i] = sum(StormData[StormData$YEARY==y,]$PROPDMGC)
  i <- i+1
}
plot.df <- data.frame(unique.years, Damage)
colnames(plot.df) <- c("year", "Damage")
# Now create a ggplot with the data frame
print(
  plot.df %>%
    gather(Legend,Count, Damage) %>%
    ggplot(aes(x=year,y=Count,color=Legend)) + geom_point() +
    labs(x="Year", y="Property Damage in US Dollars") +
    ggtitle("Property Damage Due to Storm Events in the United States"))
```

The following plot shows cost damage across the United States for `r worst.c.name` events.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
Damage <- numeric(length(unique.years))
i <- 1
for( y in unique.years ) {
  Damage[i] = sum(StormData[StormData$YEARY==y,]$CROPDMGC)
  i <- i+1
}
#print(Damage)
plot.df <- data.frame(unique.years, Damage)
colnames(plot.df) <- c("year", "Damage")
# Now create a ggplot with the data frame
print(
  plot.df %>%
    gather(Legend,Count, Damage) %>%
    ggplot(aes(x=year,y=Count,color=Legend)) + geom_point() +
    labs(x="Year", y="Crop Damage in US Dollars") +
    ggtitle("Crop Damage Due to Storm Events in the United States"))
```

## Body and Property Damage by State

States and their executive and legistlative bodies are responsible for risk assessment with respect to
weather, storm, and other such events.
Breaking down both life and
property events to each state is helpful for state governments to understand what types of
events are likely to occur in their states and what the potential risks and costs are
for risk assessment planning.  Breaking these values also can help the United States Congress
to plan the budget to set aside funds for helping states in times of emergency.

```{r eval=TRUE, echo=TRUE, cache=TRUE, results="asis"}
# Extract data per state into its own frame
for(st in sort(unique(StormData$STATE))) {
  state.report <- StormData[StormData$STATE==st,]
# Then, get gather body and property damage for each "state"
  fatalities <- injuries <- properties <- crops <-
    vector(mode="numeric", length=length(unique(state.report$EVTYPE)))
  i <- 1
  wkevt <- unique(state.report$EVTYPE)
  for(ev in wkevt) {
    fatalities[i] <- sum(state.report[state.report$EVTYPE==ev,]$FATALITIES)
    injuries[i] <- sum(state.report[state.report$EVTYPE==ev,]$INJURIES)
    properties[i] <- sum(state.report[state.report$EVTYPE==ev,]$PROPDMGC)
    crops[i] <- sum(state.report[state.report$EVTYPE==ev,]$CROPDMGC)
    i <- i+1
  }
  names(fatalities) <- names(injuries) <- names(properties) <- names(crops) <- wkevt
  # Sort each
  fatalities <- sort(fatalities, decreasing=TRUE)
  injuries <- sort(injuries, decreasing=TRUE)
  properties <- sort(properties, decreasing=TRUE)
  crops <- sort(crops, decreasing=TRUE)
  # Create a nice report for each "state"
  cat("##", st, "\n")
  cat("###", "Events of Concern", "\n")
  cat(sprintf("%s  %s", 
      "Top events for fatalities, injuries, property damage and crop damage (up to three).",
      "The data includes all recorded weather events from 1950 to November 2011."))
  
  mnrow <- ifelse(length(fatalities) < 3, length(fatalities), 3)
  df <- data.frame(c(prettyNum(unlist(fatalities[1:mnrow]),scientific=FALSE,big.mark=",")))
  colnames(df) <- c("Number of Fatalities")
  print(knitr::kable(df, caption="Fatalities (1950-2011)"))
  cat("\n")
  
  mnrow <- ifelse(length(injuries) < 3, length(injuries), 3)
  df <- data.frame(c(prettyNum(unlist(injuries[1:mnrow]),scientific=FALSE,big.mark=",")))
  colnames(df) <- c("Number of Injuries")
  print(knitr::kable(df, caption="Injuries (1950-2011)"))
  cat("\n")
  
  mnrow <- ifelse(length(properties) < 3, length(properties), 3)
  df <- data.frame(c(prettyNum(unlist(properties[1:mnrow]),scientific=FALSE,big.mark=",")))
  colnames(df) <- c("Property Damage ($)")
  print(knitr::kable(df, caption="Property Damage Costs (1950-2011)"))
  cat("\n")
  
  mnrow <- ifelse(length(crops) < 3, length(crops), 3)
  df <- data.frame(c(prettyNum(unlist(crops[1:mnrow]),scientific=FALSE,big.mark=",")))
  colnames(df) <- c("Crop Damage ($)")
  print(knitr::kable(df, caption="Crop Damage Costs (1950-2011)"))
  cat("\n")

}
```
